Statistical Reporting Inconsistencies in Experimental Linguistics

### 2024-07-25 manuscript/manuscript_anonymized.tex
anonymize

\authorsnames{Dara Leonard Jenssen Etemady,Timo B. Roettger}

\affiliation{
{Department of Linguistics \& Scandinavian Studies, University of Oslo}}

\leftheader{Etemady and Roettger}

\abstract{The present paper investigates the prevalence of statistical
reporting consistencies across articles in eight experimental
linguistics journals, published between 2000 and 2023. Using the R
package ``Statcheck'', we retrieved 39532 statistical test and assessed
whether p-values were inconsistent with their test statistic and degrees
of freedom. Half of all articles that used null hypothesis signficance
testing contained at least one inconsistent p-value. One in eight
articles contained an inconsistency that may have affected the
statistical conclusion. The inconsistency rates were stable across
journals and publication years. We offer actionable steps for authors,
reviewers, and editors to remmedy this state-of-affairs.}

\keywords{statistics, statcheck, reproducibility}

\authornote{\par{\addORCIDlink{Timo B. Roettger}{0000-0000-0000-0001}} 
\par{ }
\par{   The authors have no conflict of interest to
declare.   Conceptionalization, Methodology, Validation, Formal
Analysis, Review \& Editing of Manuscript, Data Curation - T.B.R. \&
D.L.J.E.; Software, Investigation - D.L.J.E.; Writing of Original Draft,
Visualization, Supervision - T.B.R. }
\par{Correspondence concerning this article should be addressed to Timo
B. Roettger, Email: timo.roettger@gmail.com}
}

@@

What we know about human language and its cognitive underpinnings is
often informed by experimental data. Researchers test theoretical
predictions with these data using statistical tests. Depending on the
results of such tests, researchers make claims for or against
theoretical assumptions.\\
Since they play such an integral part in the reasoning process of
experimentalists, both the data these tests are based on and the
computational procedure of the tests itself should be transparent in
order for other researchers to critically evaluate them. Moreover things
can go wrong. To err is human. Transparent sharing allows others to
detect and correct human error. In the recent years, the quantitative
sciences have seen repeated calls to become more transparent and
reproducible by sharing data and statistical protocols
(\citeproc{ref-arvan2022reproducibility}{Arvan et al., 2022};
\citeproc{ref-laurinavichyute2022share}{Laurinavichyute et al., 2022};
\citeproc{ref-roettger2019researcher}{Roettger, 2019}). Despite these
calls, sharing of statistical protocols is still rather rare across the
language sciences (\citeproc{ref-bochynska2023reproducible}{Bochynska et
al., 2023}). If statistical procedures cannot be critically evaluated,
human errors might be left undetected and thus remain uncorrected in the
publication record. And if undetected errors affect the decision
procedure of the analysis, i.e.~whether a hypothesis is rejected or
accepted, these errors might lead to -at best- overconfident, -at worst-
false theoretical conclusions. The present paper will present evidence
that the published literature in experimental linguistics contains a
concerning amount of such statistical errors, a state of affairs which
warrants more rigorous data sharing practices.

@@

Since data and statistical protocol sharing remains rare across the
language sciences (\citeproc{ref-bochynska2023reproducible}{Bochynska et
al., 2023}), interested readers are left with trusting the authors that
the statistical analysis is run and reported correctly. However, the
three sets of indices in (1) have clearly defined mathematical
relationships and can thus be easily checked for consistency. An F test
with the specified degrees of freedom and a test-statistic of 3.88
should result in a p-value of 0.053 which is larger, not smaller, than
0.05. Possible reasons for this inconsistency are manifold: It could be
a typo of the comparison sign, i.e.~the authors meant to use \texttt{=}
or \texttt{\textgreater{}} rather than \texttt{\textless{}}. Any of the
numbers could contain a typo. Sometimes an error might indicate
erroneous rounding (e.g.~0.057 being rounded down to 0.05). Without
access to data and scripts, it remains unclear to the reader what has
caused this inconsistency. Such inconsistencies can be particularly
concerning if the calculated p-value (here 0.057) and the reported
p-values (here \textless0.05) are not on the same side of the alpha
threshold. In NHST, p-values below a conventionalized alpha threshold,
most commonly 0.05, are interpreted as evidence that the data are
sufficiently inconsistent with the null hypothesis (``significant'').
P-values above that threshold are considered consistent with the null
hypothesis and practically lead to rejecting the alternative hypothesis
(``non-significant''). In (1) above, the reported p-value suggest a
significant result, the p-value derived from the degrees of freedom and
the test statistic suggest a non-significant result. In the following we
refer to these inconsistencies as ``decision inconsistencies''.

The consistency of these values can be automatically assessed if
statistical tests are reported in an unambiguous format. Recently, a
series of studies used such automatic assessments to evaluate the
prevalence of inconsistent statistical reporting in psychology
(\citeproc{ref-bakker2011mis}{Bakker \& Wicherts, 2011},
\citeproc{ref-bakker2014outlier}{2014};
\citeproc{ref-caperos2013consistency}{Caperos \& Pardo, 2013};
\citeproc{ref-claesen2023data}{Claesen et al., 2023};
\citeproc{ref-green2018statcheck}{Green et al., 2018};
\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016};
\citeproc{ref-nuijten2020statcheck}{Nuijten \& Polanin, 2020};
\citeproc{ref-veldkamp2014statistical}{Veldkamp et al., 2014};
\citeproc{ref-wicherts2011willingness}{Wicherts et al., 2011}), medical
sciences (\citeproc{ref-garcia2004incongruence}{Garcı́a-Berthou \&
Alcaraz, 2004}; \citeproc{ref-van2023comparing}{Van Aert et al., 2023}),
psychiatry (\citeproc{ref-berle2007inconsistencies}{Berle \& Starcevic,
2007}), cyber security studies (\citeproc{ref-gross2021fidelity}{Groß,
2021}), technological education research
(\citeproc{ref-buckley2023estimating}{Buckley et al., 2023}), and
experimental philosophy (\citeproc{ref-colombo2018statistical}{Colombo
et al., 2018}). For example, looking at over 250 thousand p-values
published in major psychology journals, Nuijten et al.
(\citeproc{ref-nuijten2016prevalence}{2016}) found that around 50\% of
the articles with statistical results contained at least one
inconsistency and around 13\% contained at least one ``decision
inconsistency''. Other studies report on inconsistency rates between 4\%
and 14\%, with between 10\% and 63\% of articles containing at least one
inconsistency and between 3\% and 21\% decision consistencies.

To assess the prevalence of statistical-reporting inconsistencies in
experimental linguistics, the present paper conceptually replicates
Nuijten et al. (\citeproc{ref-nuijten2016prevalence}{2016}) and assesses
p-values reported in eight experimental linguistic journals published
between 2000 and 2023. We explore whether the rate of inconsistency
differs across journals, whether that rate has changed over the course
of the last 20 years and whether there is evidence for bias in these
statistical-reporting inconsistencies. We discuss the results and offer
concrete recommendations for authors, reviewers, and editors to tackle
this problem.

@@

All quantitative analyses were conducted using R Core Team
(\citeproc{ref-rmanual}{2021}) and the r packages {[}LIST ALL PACKAGES
from 01\_Analysis.R and add to references{]}

\subsection{3.1. Statcheck}\label{statcheck}

We used the R package Statcheck (Version 1.4.1-beta.2,
\citeproc{ref-nuijten2023statcheck}{Nuijten \& Epskamp, 2023}) to
automatically detect statistical-reporting inconsistencies. Statcheck
works as follows: After converting articles in pdf or html format to
plain text, Statcheck searches for specific strings that correspond to a
NHST result using ``regular expressions''. That way, statcheck can
detect results of t-tests, F-tests, Z-tests, χ2-tests, correlation
tests, and Q tests as long as the test result fulfills three conditions:
(a) the test result is reported completely including the test statistic,
the degrees of freedom (if applicable), and the p-value; (b) the test
result is in the body of the text, i.e.~Statcheck usually misses
reported in tables; and (c) the test result is reported in American
Psychological Association style (\citeproc{ref-APA2020}{American
Psychological Association, 2020}). Given these constraints, Statcheck is
estimated to detect roughly 60\% of all reported NHST results
(\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}). Statcheck
uses the reported test statistic and degrees of freedom to recalculate
the p-value, compares the reported and recalculated p-values and, if
there is a mismatch, flags the test as containing an ``inconsistency.''
The algorithm takes into account that tests might have been performed as
one-tailed by identifying the search strings ``one-tailed,''
``one-sided,'' or ``directional'' in the body of the text. Moreover,
Statcheck considers p = .000 and p \textless{} .000 as inconsistent
because p-values of exactly zero are mathematically impossible and the
APA manual (\citeproc{ref-APA2020}{American Psychological Association,
2020}) advises to report very small p-values as p \textless{} .001.
Validity checks of Statcheck suggest that inter-rater reliability
between manual coding and Statcheck is high, i.e.~0.76 for
inconsistencies and 0.89 for decision inconsistencies
(\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}). The
overall accuracy of Statcheck is estimated to be between 96.2\% to
99.9\% (\citeproc{ref-nuijten2017validity}{Nuijten et al., 2017}). We
thus consider Statcheck a valid proxy of the prevalence of statistical
reporting inconsistencies.

\subsection{3.2. Sample}\label{sample}

@@

\subsection{3.3. Data availability
statement}\label{data-availability-statement}

All derived data and corresponding R scripts are publicly available
here: LINK.

@@

When examining reported against recalculated p-values for all
inconsistencies (see Figure~\ref{fig-scatter}), we can identify certain
spatial patterns (only inconsistencies that are based on `=' comparisons
were included to make them easier to interpret, n = 3779 out of 5166).
First, the center of density of points is located in the bottom left
corner with more p-values reported closer to the standard alpha level of
0.05. This is not surprising, since the quantitative sciences exhibit a
documented publication bias, with hypotheses being more often confirmed
(and thus p \textless{} .05) than not
(\citeproc{ref-franco2014publication}{Franco et al., 2014};
\citeproc{ref-sterling1959publication}{Sterling, 1959}). Second, there
is a linear density of points along the diagonal axis which corresponds
to numerically small inconsistencies, some of which might be related to
simple rounding errors. However, comparing the diagonal to the black
line, which represents a linear model predicting recalculated by
reported p-values, we can see a clear divergence of what is expected if
inconsistencies were equally likely in both directions. The slope of the
regression line is flatter than the diagonal axis which means that, on
average, inconsistencies have a tendency to exhibit smaller reported
p-values than their recalculated counterparts. Similarly,
inconsistencies that report the p-value as being larger or smaller than
a reference value (e.g.~p \textgreater{} 0.05 or p \textless{} 0.05 )
are not equally prevalent. There were 4.4\% of inconsistencies with p
being reported as larger than a reference but 7.4\% inconsistencies with
p being reported as smaller than a reference (e.g.~p \textless{} 0.05).
So even if we assumed these inconsistencies were merely typos of the
comparison sign, inconsistencies that erroneously report the p-value to
be smaller than a reference value are more frequent than inconsistencies
that erroneously report the p-value to be larger than a reference value.
These biases are also reflected in decision inconsistencies. Of all
decision inconsistencies (n = 655), 71\% represent cases in which a
reported significant result (p \textless{} 0.05) is recalculated as
non-significant (p \textgreater{} 0.05).

@@

The present study found a large amount of statistical reporting
inconsistencies across a sample of 5804 experimental linguistic
articles, containing 3059 assessable p-values. 13.1\% of all p-values
were flagged as inconsistent and 1.7\% were flagged as decision
inconsistencies, i.e.~the reported p-value is on the opposite side of
the alpha threshold than the recalculated p-value. On average, 50\% of
assessable articles contained at least one inconsistency and 13\%
contained at least one decision inconsistency.

The present study can be considered a conceptual replication of previous
studies investigating statistical-reporting inconsistency across
different disciplines (\citeproc{ref-bakker2011mis}{Bakker \& Wicherts,
2011}, \citeproc{ref-bakker2014outlier}{2014};
\citeproc{ref-caperos2013consistency}{Caperos \& Pardo, 2013};
\citeproc{ref-veldkamp2014statistical}{Veldkamp et al., 2014};
\citeproc{ref-wicherts2011willingness}{Wicherts et al., 2011}) and most
recent such assessment using Statcheck
(\citeproc{ref-buckley2023estimating}{Buckley et al., 2023};
\citeproc{ref-colombo2018statistical}{Colombo et al., 2018};
\citeproc{ref-gross2021fidelity}{Groß, 2021};
\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}). Previous
studies report on inconsistency rates between 4\% and 14\%, with between
10\% and 63\% of articles containing at least one inconsistency and
between 3\% and 21\% contain at least one decision consistency.

Even if the prevalence of these inconsistencies could be largely
attributed to inconsequential typos or rounding errors (an assumption we
cannot test without access to the data), the sheer amount of these
inconsistencies that have made it through peer-review should concern us.
They are human errors. If such a substantial amount of errors is found
in plain site, the question naturally arises as to how many errors
during the data analysis itself remain undetected. If the tip of the
iceberg is already so large, what is the volume of the submerged
iceberg?

The present examination assessment did not indicate any noticeable trend
over the last 23 years of publications nor did the data suggest large
differences between journals.

Observed inconsistencies were characterized by reported p-values being
on average lower than their recalculated counterparts and the prevalence
of decision inconsistencies was higher for p-values reported as
significant than for those reported as non-significant. These patterns
could indicate a systematic bias in favor of lower p-values in general
and a bias towards significant results in particular. Our data do not
speak to the causes of these biases, but possible reasons include the
following: Researchers might intentionally round down p-values because
they think a lower p-value is more convincing to reviewers and/or
readers. This practice has been admitted to by 1 in 5 surveyed
psychological researchers (\citeproc{ref-john2012measuring}{John et al.,
2012}). Given that a non-trivial amount of quantitative linguists have
admitted to commit questionable research practices (and even fraud)
(\citeproc{ref-isbell2022misconduct}{Isbell et al., 2022}), we cannot
exclude the possibility that some of the inconsistencies were
intentional. It is our strong belief, however, that the majority of
inconsistencies are unintentional.

Researchers might scrutinize non-significant results more than
significant results or are less likely to double check significant
results than non-significant results because results that confirm their
hypothesis feed into their confirmation bias
(\citeproc{ref-nickerson1998confirmation}{Nickerson, 1998}). For
example, Fugelsang et al. (\citeproc{ref-fugelsang2004theory}{2004}) let
researchers evaluate data that are either consistent or inconsistent
with their prior expectations. They showed that when researchers
encounter results that disconfirm their expectations, they are likely to
blame the methodology while results that confirmed their expectations
were rarely critically scrutinized. Alternatively, the observed bias
might be a reflection of publication bias
(\citeproc{ref-franco2014publication}{Franco et al., 2014};
\citeproc{ref-sterling1959publication}{Sterling, 1959}) with
(erroneously) reported significant p-values being more likely to be
published than non-significant ones.

@@

While we believe our work offers an important contribution to improving
statistical reporting practices in experimental linguistics, there are,
of course, a number of limitations. The present assessment and the
conclusions we can draw from them are limited. First, our sample is
limited to only a subset of experimental linguistic journals. Our sample
is based on a crude criterion of what constitutes an experimental
linguistic journal (see \citeproc{ref-kobrock2023assessing}{Kobrock \&
Roettger, 2023}) and we restricted ourselves further to (for us)
accessible journals which explicitly require APA statistical formatting
in the author guidelines. Thus it is possible that a different selection
of journals would have resulted in different results. However, given
that the inconsistency rates of our study are comparable to similar
studies from other disciplines and that those rates are relatively
stable across journals and time, our findings should be considered
relevant for experimental linguistics at large.

@@

There are concrete actionable steps the field of experimental
linguistics can make to reduce statistical reporting inconsistencies. In
order to avoid simple copy-and-paste errors related to working in two
separate programs for writing the manuscript and conducting the
statistical analysis, authors should consider `literate programming',
i.e.~an integration of analysis code and prose into a single, dynamic
document (\citeproc{ref-casillas2023opening}{Casillas et al., 2023};
\citeproc{ref-knuth1984literate}{Knuth, 1984}). Several implementations
of literate programming are freely available to researchers including
common R markdown files (Rmd) and Quarto markdown files (qmd). Literate
programming can ensure that values derived from the statistical analysis
are automatically integrated into the manuscript document, avoiding
errors that might happen during a manual transfer from one program to
the other.

@@

Journal editors could explicitly recommend consistency checks with
algorhythms such as statcheck during peer review, a practice that has
been taken up on by several journals from neighboring disciplines
{[}Psychological Science\footnote{http://www.psychologicalscience.org/publications/psychological\_science/ps-submissions;
  accessed on July 15, 2024.}, Advances in Methods and Practices in
Psychological Science\footnote{https://www.psychologicalscience.org/publications/ampps/ampps-submission-guidelines;
  accessed on on July 15, 2024.}, Stress \& Health Barber
(\citeproc{ref-barber2017meticulous}{2017}){]}. Editors could also
demand, recommend or at least encourage data sharing for publication in
their journal. Data sharing policies have been shown to substantially
increase the reproducibility of analyses
(\citeproc{ref-hardwicke2018data}{Hardwicke et al., 2018}; e.g.
\citeproc{ref-laurinavichyute2022share}{Laurinavichyute et al., 2022}).

>>>>

\author{anonymous}

\affiliation{
{anonymous}}

\leftheader{anonymous}

\abstract{The present paper investigates the prevalence of statistical
reporting consistencies across articles in eight experimental
linguistics journals published between 2000 and 2023. Using the R
package ``Statcheck'', we retrieved 39532 statistical test and assessed
whether p-values were consistent with their test statistic and degrees
of freedom. Half of all articles that used null hypothesis signficance
testing contained at least one inconsistent p-value. One in eight
articles contained an inconsistency that may have affected the
statistical conclusion. The inconsistency rates were comparable across
journals and seem stable over publication years. We discuss possible
reasons for this high rate and offer actionable steps for authors,
reviewers, and editors to remedy this state-of-affairs.}

\keywords{statistics, statcheck, reproducibility, null hypothesis
significance testing}

\authornote{ 
\par{ }
\par{   The authors have no conflict of interest to declare.    }
\par{Correspondence concerning this article should be addressed to }
}

@@

What we know about human language and its cognitive underpinnings is
often informed by experimental data. Researchers test theoretical
predictions with their data using statistical tests. Depending on the
results of their tests, researchers make claims for or against
theoretical assumptions.\\
Since these tests play such an integral part in the argumentation
process of experimentalists, both the data these tests are based on and
the computational procedure of the tests itself should be transparent in
order for other researchers to critically evaluate them. But more than
that, things can go wrong. Humans are fallible; they make mistakes.
Transparent sharing allows others to detect and correct human error. In
the recent years, quantitative linguistics have seen repeated calls to
become more transparent and reproducible by sharing data and statistical
protocols (\citeproc{ref-arvan2022reproducibility}{Arvan et al., 2022};
\citeproc{ref-laurinavichyute2022share}{Laurinavichyute et al., 2022};
\citeproc{ref-roettger2019researcher}{Roettger, 2019}). Despite these
calls, sharing of statistical protocols is still rather rare across the
language sciences (\citeproc{ref-bochynska2023reproducible}{Bochynska et
al., 2023}). If statistical procedures cannot be critically evaluated,
human errors might be left undetected and thus remain uncorrected in the
publication record. And if undetected errors affect the decision
procedure of the analysis, i.e.~whether a hypothesis is rejected or
accepted, these errors might lead to -at best- overconfident, -at worst-
false theoretical conclusions. The present paper will present evidence
that the published literature in experimental linguistics contains a
concerning amount of such statistical errors, a state of affairs which
warrants more rigorous data sharing practices.

@@

Without access to the data and scripts, interested readers are left with
trusting the authors that the statistical analysis is run and reported
correctly. However, the three sets of indices in (1) have clearly
defined mathematical relationships and can thus be easily assessed for
consistency. For example, an F test with the specified degrees of
freedom and a test-statistic of 3.88 should result in a p-value of 0.053
which is larger, not smaller, than 0.05. Possible reasons for this
inconsistency are manifold: It could be a typo of the comparison sign,
i.e.~the authors meant to use \texttt{=} or \texttt{\textless{}} rather
than \texttt{\textgreater{}}. Additionally, any of the numbers could
contain a typo and sometimes an error might indicate erroneous rounding
(e.g.~0.057 being rounded down to 0.05). Without access to data and
scripts, it remains unclear to the reader what has caused this
inconsistency. Such inconsistencies can be particularly concerning if
the calculated p-value (here 0.057) and the reported p-values (here
\textless0.05) are not on the same side of the alpha threshold. In NHST,
p-values below a conventionalized alpha threshold, most commonly 0.05,
are interpreted as evidence that the data are sufficiently inconsistent
with the null hypothesis (``significant''). P-values above that
threshold are considered consistent with the null hypothesis and
practically lead to rejecting the alternative hypothesis
(``non-significant''). In (1) above, the reported p-value suggest a
significant result, the p-value derived from the degrees of freedom and
the test statistic suggest a non-significant result. In the following we
refer to these inconsistencies as ``decision inconsistencies''.

The consistency of these values can be automatically assessed if
statistical tests are reported in an unambiguous format. Recently, a
series of studies used such automatic assessments to evaluate the
prevalence of inconsistent statistical reporting in psychology
(\citeproc{ref-bakker2011mis}{Bakker \& Wicherts, 2011},
\citeproc{ref-bakker2014outlier}{2014};
\citeproc{ref-caperos2013consistency}{Caperos \& Pardo, 2013};
\citeproc{ref-claesen2023data}{Claesen et al., 2023};
\citeproc{ref-green2018statcheck}{Green et al., 2018};
\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016};
\citeproc{ref-nuijten2020statcheck}{Nuijten \& Polanin, 2020};
\citeproc{ref-veldkamp2014statistical}{Veldkamp et al., 2014};
\citeproc{ref-wicherts2011willingness}{Wicherts et al., 2011}), medical
sciences (\citeproc{ref-garcia2004incongruence}{Garcı́a-Berthou \&
Alcaraz, 2004}; \citeproc{ref-van2023comparing}{Van Aert et al., 2023}),
psychiatry (\citeproc{ref-berle2007inconsistencies}{Berle \& Starcevic,
2007}), cyber security studies (\citeproc{ref-gross2021fidelity}{Groß,
2021}), technological education research
(\citeproc{ref-buckley2023estimating}{Buckley et al., 2023}), and
experimental philosophy (\citeproc{ref-colombo2018statistical}{Colombo
et al., 2018}). For example, looking at over 250 thousand p-values
published in major psychology journals, Nuijten et al.
(\citeproc{ref-nuijten2016prevalence}{2016}) found that around 50\% of
the articles with statistical results contained at least one
inconsistency and around 13\% contained at least one decision
inconsistency. Other studies report on inconsistency rates between 4\%
and 14\%, with between 10\% and 63\% of articles containing at least one
inconsistency and between 3\% and 21\% contain at least one decision
consistency. These high inconsistency rates in other disciplines are
concerning and have led to a constructive dialog, resulting in
recommendations for best practice to either avoid inconsistencies or
make them more detectable.

To assess the prevalence of statistical-reporting inconsistencies in
experimental linguistics, the present paper conceptually replicates
Nuijten et al. (\citeproc{ref-nuijten2016prevalence}{2016}) and assesses
p-values reported in eight experimental linguistic journals published
between 2000 and 2023. We explore whether the rate of inconsistency
differs across journals, whether their rate has changed over the course
of the last 23 years and whether there is evidence for bias in these
statistical-reporting inconsistencies. We discuss the results and offer
concrete recommendations for authors, reviewers, and editors to tackle
this problem.

@@

\subsection{3.1. Data availability
statement}\label{data-availability-statement}

All quantitative analyses were conducted using R Core Team
(\citeproc{ref-rmanual}{2021}). All derived data and corresponding R
scripts are publicly available here:
\url{https://github.com/timo-b-roettger/ErrorsLing}

\subsection{3.2. Statcheck}\label{statcheck}

We used the R package Statcheck (Version 1.4.1-beta.2,
\citeproc{ref-nuijten2023statcheck}{Nuijten \& Epskamp, 2023}) to
automatically detect statistical-reporting inconsistencies. Statcheck
works as follows: After converting articles in pdf or html format to
plain text, Statcheck searches for specific strings that correspond to a
NHST result using regular expressions. That way, Statcheck can detect
results of t-tests, F-tests, Z-tests, χ2-tests, correlation tests, and Q
tests as long as the test result fulfills three conditions: (a) the test
result is reported completely including the test statistic, the degrees
of freedom (if applicable), and the p-value; (b) the test result is in
the body of the text, i.e. Statcheck usually misses information in
tables; and (c) the test result is reported in American Psychological
Association style (\citeproc{ref-APA2020}{American Psychological
Association, 2020}). Given these constraints, Statcheck is estimated to
detect roughly 60\% of all reported NHST results
(\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}). Statcheck
uses the reported test statistic and degrees of freedom to recalculate
the p-value, compares the reported and recalculated p-value and, if
there is a mismatch, flags the test as containing an ``inconsistency.''
The algorithm takes into account that tests might have been performed as
one-tailed by identifying the search strings ``one-tailed,''
``one-sided,'' or ``directional'' in the body of the text. Moreover,
Statcheck considers p = .000 and p \textless{} .000 as inconsistent
because p-values of exactly zero are mathematically impossible and the
APA manual (\citeproc{ref-APA2020}{American Psychological Association,
2020}) advises to report very small p-values as p \textless{} .001.
Validity checks of Statcheck suggest that inter-rater reliability
between manual coding and Statcheck is high, i.e.~0.76 for
inconsistencies and 0.89 for decision inconsistencies
(\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}). The
overall accuracy of Statcheck is estimated to be between 96.2\% to
99.9\% (\citeproc{ref-nuijten2017validity}{Nuijten et al., 2017}). We
thus consider Statcheck a valid proxy of the prevalence of statistical
reporting inconsistencies.

\subsection{3.3. Sample}\label{sample}

@@

@@

When examining reported against recalculated p-values for all
inconsistencies (see Figure~\ref{fig-scatter}), we can identify certain
spatial patterns (only inconsistencies that are based on `=' comparisons
were included to make them easier to interpret, n = 3779 out of 5166).
First, the center of density of points is located in the bottom left
corner with more p-values reported closer to the most common alpha level
of 0.05. This is not surprising, since the quantitative sciences exhibit
a documented publication bias, with hypotheses being more often
confirmed (and thus p \textless{} .05) than not
(\citeproc{ref-franco2014publication}{Franco et al., 2014};
\citeproc{ref-sterling1959publication}{Sterling, 1959}). Second, there
is a linear density of points along the diagonal axis which corresponds
to numerically small inconsistencies, some of which might be related to
simple rounding errors. However, comparing the diagonal to the black
line, which represents a linear model predicting recalculated by
reported p-values, we can see a clear divergence of what is expected if
inconsistencies were equally likely in both directions. The slope of the
regression line is flatter than the diagonal axis which means that, on
average, inconsistencies have a tendency to exhibit smaller reported
p-values than their recalculated counterparts. Similarly,
inconsistencies that report the p-value as being larger or smaller than
a reference value (e.g.~p \textgreater{} 0.05 or p \textless{} 0.05 )
are not equally prevalent. There were 4.4\% of inconsistencies with p
being reported as larger than a reference but 7.4\% inconsistencies with
p being reported as smaller than a reference (e.g.~p \textless{} 0.05).
So even if we assumed these inconsistencies were merely typos of the
comparison sign, inconsistencies that erroneously report the p-value to
be smaller than a reference value are more frequent than inconsistencies
that erroneously report the p-value to be larger than a reference value.
These biases are also reflected in decision inconsistencies. Of all
decision inconsistencies (n = 655), 71\% represent cases in which a
reported significant result (p \textless{} 0.05) is recalculated as
non-significant (p \textgreater{} 0.05).

@@

The present study found a large amount of statistical reporting
inconsistencies across a sample of 5804 experimental linguistic
articles, containing 39532 assessable p-values. 13.1\% of all p-values
were flagged as inconsistent and 1.7\% were flagged as decision
inconsistencies, i.e.~the reported p-value is on the opposite side of
the alpha threshold than the recalculated p-value. On average, 50\% of
assessable articles contained at least one inconsistency and 13\%
contained at least one decision inconsistency.

The present study can be considered a conceptual replication of previous
studies investigating statistical-reporting inconsistency across
different disciplines (\citeproc{ref-bakker2011mis}{Bakker \& Wicherts,
2011}, \citeproc{ref-bakker2014outlier}{2014};
\citeproc{ref-caperos2013consistency}{Caperos \& Pardo, 2013};
\citeproc{ref-veldkamp2014statistical}{Veldkamp et al., 2014};
\citeproc{ref-wicherts2011willingness}{Wicherts et al., 2011}) and most
recent such assessment using Statcheck
(\citeproc{ref-buckley2023estimating}{Buckley et al., 2023};
\citeproc{ref-colombo2018statistical}{Colombo et al., 2018};
\citeproc{ref-gross2021fidelity}{Groß, 2021};
\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}). The
discovered rates fall in line with these studies who report on
inconsistency rates between 4\% and 14\%, with between 10\% and 63\% of
articles containing at least one inconsistency.

Even if the prevalence of these inconsistencies could be largely
attributed to inconsequential typos or rounding errors (an assumption we
cannot test without access to the data), the sheer amount of these
inconsistencies that have made it through peer-review should concern us.
They are human errors. If such a substantial amount of errors is found
in plain site, the question naturally arises as to how many errors
during the data analysis itself remain undetected. We should ask
ourselves, if the tip of the iceberg is already so large, what is the
volume of the submerged iceberg?

The present examination did not indicate any noticeable trend over the
last 23 years of publications. Observed inconsistencies were
characterized by reported p-values being on average lower than their
recalculated counterparts and the prevalence of decision inconsistencies
was higher for p-values reported as significant than for those reported
as non-significant. These biases have been reported on for other
disciplines as well (\citeproc{ref-bakker2011mis}{Bakker \& Wicherts,
2011}; \citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}).
These skewed patterns could indicate a systematic bias in favor of lower
p-values in general and a bias towards significant results in
particular. Our data do not speak to the causes of these biases, but
possible reasons include the following: Researchers might intentionally
round down p-values because they think lower p-values are more
convincing to reviewers and readers. This practice has been admitted to
by 1 in 5 surveyed psychological researchers
(\citeproc{ref-john2012measuring}{John et al., 2012}). Given that a
non-trivial amount of quantitative linguists have admitted to commit
questionable research practices (and even fraud)
(\citeproc{ref-isbell2022misconduct}{Isbell et al., 2022}), we cannot
exclude the possibility that some of the inconsistencies in our sample
were indeed intentional. It is our strong belief, however, that the
majority of inconsistencies are unintentional.

Researchers might scrutinize non-significant results more than
significant results or are less likely to double check significant
results than non-significant results because results that confirm their
hypothesis feed into their confirmation bias
(\citeproc{ref-nickerson1998confirmation}{Nickerson, 1998}). For
example, Fugelsang et al. (\citeproc{ref-fugelsang2004theory}{2004}) let
researchers evaluate data that are either consistent or inconsistent
with their prior expectations. They showed that when researchers
encounter results that are not in line with their expectations, they are
likely to blame the methodology while results that confirmed their
expectations were rarely critically scrutinized. Alternatively, the
observed bias might merely be a reflection of publication bias
(\citeproc{ref-franco2014publication}{Franco et al., 2014};
\citeproc{ref-sterling1959publication}{Sterling, 1959}) with
(erroneously) reported significant p-values being more likely to be
published than non-significant ones.

@@

While we believe our work offers an important contribution to improving
statistical reporting practices in experimental linguistics, the present
assessment and the conclusions we can draw from them are limited: First,
our sample is limited to only a subset of experimental linguistic
journals. Our sample is based on a crude criterion of what constitutes
an experimental linguistic journal (see
\citeproc{ref-kobrock2023assessing}{Kobrock \& Roettger, 2023}) and we
restricted ourselves further to (for us) accessible journals which
explicitly require APA statistical formatting in the author guidelines.
Thus it is possible that a different selection of journals would have
resulted in different results. However, given that the inconsistency
rates of our study are comparable to similar studies from other
disciplines and that those rates are relatively stable across journals
and time, our findings should be considered relevant for experimental
linguistics at large.

@@

There are concrete actionable steps the field of experimental
linguistics can take to reduce statistical reporting inconsistencies. In
order to avoid simple copy-and-paste errors related to working in two
separate programs for writing the manuscript and conducting the
statistical analysis, authors should consider `literate programming',
i.e.~an integration of analysis code and prose into a single, dynamic
document (\citeproc{ref-casillas2023opening}{Casillas et al., 2023};
\citeproc{ref-knuth1984literate}{Knuth, 1984}). Several implementations
of literate programming are freely available to researchers including
common R markdown files (Rmd) and Quarto markdown files (qmd). Literate
programming can ensure that values derived from the statistical analysis
are automatically integrated into the manuscript document, avoiding
errors that might happen during a manual transfer from one program to
the other.

@@

Journal editors could explicitly recommend consistency checks with
algorhythms such as Statcheck during peer review, a practice that has
been taken up on by several journals from neighboring disciplines
(Psychological Science\footnote{http://www.psychologicalscience.org/publications/psychological\_science/ps-submissions;
  accessed on July 15, 2024.}, Advances in Methods and Practices in
Psychological Science\footnote{https://www.psychologicalscience.org/publications/ampps/ampps-submission-guidelines;
  accessed on on July 15, 2024.}, Stress \& Health Barber
(\citeproc{ref-barber2017meticulous}{2017})). Editors could also demand,
recommend or at least encourage data sharing for publication in their
journal. Data sharing policies have been shown to substantially increase
the reproducibility of analyses (e.g.,
\citeproc{ref-hardwicke2018data}{Hardwicke et al., 2018};
\citeproc{ref-laurinavichyute2022share}{Laurinavichyute et al., 2022}).
